---
title: "Getting Started with excel2r"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with excel2r}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

**excel2r** converts Excel formula workbooks into standalone R scripts.
It extracts every formula, translates 62 Excel functions to R equivalents,
resolves cross-sheet references, determines execution order via topological
sort, and optionally exports raw data as tidy CSVs so the `.xlsx` file is
no longer needed.

## Installation

Install from GitHub:

```{r}
# install.packages("remotes")
remotes::install_github("emantzoo/excel2r")
```

## Quick Start

### One-step migration

The simplest way to migrate an Excel workbook:

```{r}
library(excel2r)

excel2r::migrate("workbook.xlsx", "output/")
```
```
#> Script written: output/generated_script.R
#> CSV data exported: output/data/ (3 sheets)
#> Verification: 42/45 match, 0 mismatches, 2 precision diffs, 1 harmless NA/error
#> Done: 45 formulas (42 ok, 2 warnings, 1 errors) across 3 sheets
```

This creates:

- `output/generated_script.R` -- a standalone R script reproducing all formulas
- `output/data/*.csv` -- tidy long-format CSVs with raw cell data for each sheet

### Process without writing files

If you want to inspect the generated script before saving:

```{r}
result <- excel2r::process("workbook.xlsx")

# View the generated R code
cat(result$script)

# Inspect the formula report
head(result$report)
```

The report data frame has one row per formula cell with columns:
`Sheet`, `Cell`, `Formula`, `R_Code`, `Status`, and `Warnings`.

### Verify against Excel

Compare R-computed values to Excel's cached results:

```{r}
result <- excel2r::process("workbook.xlsx", mode = "excel")
v <- excel2r::verify("workbook.xlsx", result)

print(v$summary)
#> $total
#> [1] 45
#> $matches
#> [1] 42
#> $value_mismatches
#> [1] 0
#> $fp_precision
#> [1] 2
#> $na_mismatches
#> [1] 1

# Inspect any mismatches
View(v$mismatches)
```

## Output Modes

### CSV mode (default, standalone)

```{r}
excel2r::migrate("workbook.xlsx", "output/", mode = "csv")
```

The generated script reads from CSVs, so the Excel file is no longer needed.
This is the recommended mode for reproducible workflows.

### Excel mode

```{r}
excel2r::migrate("workbook.xlsx", "output/", mode = "excel")
```

The generated script reads directly from the `.xlsx` file using `readxl`.
Lighter output (no CSVs) but requires the Excel file at runtime.

## Supported Excel Functions

excel2r translates 62 Excel functions:

```{r, eval = TRUE}
library(excel2r)
excel2r::supported_functions()
```

Categories include: aggregation (SUM, AVERAGE, MIN, MAX, ...), conditional
(IF, IFS, IFERROR, ...), lookup (VLOOKUP, HLOOKUP, INDEX, MATCH, XLOOKUP),
math (ROUND, ABS, SQRT, ...), text (CONCATENATE, LEFT, RIGHT, MID, ...),
logical (AND, OR, NOT), and counting (COUNT, COUNTA, COUNTIF, COUNTIFS, ...).

## Interactive App

For a guided workflow with a web UI:

```{r}
excel2r::run_app()
```

This launches a 5-step Shiny app: upload, configure, preview formulas,
download script & data, and verify results.

## Options

Key arguments for `migrate()`:

| Argument | Default | Description |
|----------|---------|-------------|
| `mode` | `"csv"` | `"csv"` for standalone, `"excel"` for xlsx-dependent |
| `include_named_tables` | `TRUE` | Detect and export Excel named tables |
| `wrap_trycatch` | `TRUE` | Wrap formula cells in tryCatch for robustness |
| `include_comments` | `TRUE` | Add original Excel formulas as `# comments` |
| `verify` | `TRUE` | Run verification after migration |
